Probabilistic hierarchical modeling of exoplanet populations using incomplete catalogs

### 2013-11-21 document/ms.tex
starting description of importance sampling scheme

We start by assuming that ever target star has some $K$ orbiting planets.
Most of these planets are unobserved because they are two small to be
detectable or because of geometry.
The graphical model in \fig{gm} is a sketch of the hypothetical generative
procedure for a catalog of exoplanet observations.
The way it is shown, this model is extremely general but we will consider a
very specific form here.
In \fig{gm} the global ``hyper'' parameters are:
\begin{itemize}
\item{the parameters describing the distributions of the exoplanet population
\population, and}
\item{the parameters of a model of the catalog's exoplanet selection function
\selection.}
\end{itemize}
Similarly, the ``hidden variables'' are:
\begin{itemize}
\item{the physical parameters of the star and the stellar system
\stellar---including the stellar mass and radius, a measure of the stellar
variability, and the mean inclination of the orbiting exoplanets---and}
\item{the physical parameters of an exoplanet \planet---period, radius, and
relative inclination.}
\end{itemize}
Finally, the observables are

>>>>

We start by assuming that ever target star has some $K$ orbiting planets.
Most of these planets are unobserved because they are two small to be
detectable or because of geometry.
The graphical model in \fig{gm} is a sketch of the hypothetical generative
procedure for a catalog of exoplanet observations.
The way it is shown, this model is extremely general but we will consider a
very specific form here.
The parameters of this model are listed and described in \tab{parameters}.

Using this model, we want to find---conditioned on the catalog of exoplanet
candidates---joint constraints on the parameters describing the population of
exoplanets and the selection function of the survey.
To this end, we need to be able to efficiently marginalize over the parameters
of all of the exoplanet systems (observed and unobserved) to compute the
marginalized likelihood function
\begin{eqnarray}
p(\planetobs,\stellarobs,\isobs\,|\,\population,\selection) &=&
\prod_{n=1}^N \int \left [
p(\stellar_n)\,p(\stellarobs_n\,|\,\stellar_n) \,
 \prod_{k=1}^K
p(\planetobs_{kn},\isobs_{kn}\,|\,\stellar_n,\population,\selection)\right ]
\dd \stellar_n
\end{eqnarray}
where
\begin{eqnarray}\eqlabel{planet-marg}
p(\planetobs_{kn},\isobs_{kn}\,|\,\stellar_n,\population,\selection) &=&
\int
p(\planet_{kn}\,|\,\population) \,
p(\planetobs_{kn}\,|\,\planet_{kn},\stellar_{n}) \,
p(\isobs_{kn}\,|\,\planet_{kn},\stellar_{n},\selection)
\dd \planet_{kn} \quad.
\end{eqnarray}

We can evaluate the integral in \eq{planet-marg} using importance sampling.
If the exoplanet is not observed ($\isobs_{kn}=0$), the most sensible proposal
density is simply $\tilde{p}(\planet_{kn})=p(\planet_{kn}\,|\,\population)$
and the integral becomes
\begin{eqnarray}
p(\planetobs_{kn},\isobs_{kn}\,|\,\stellar_n,\population,\selection) &=&
\frac{1}{\sum_l \tilde{p}(x_l)}\,\sum_l
p(\isobs_{kn}\,|\,\planet_{kn}=x_l,\stellar_{n},\selection)
\end{eqnarray}
for $x_l \sim p(\planet_{kn}\,|\,\population)$.
